Local Spectral Clustering of Density Upper Level Sets

### 2021-12-23 arxiv/ppr_clustering_neighborhood_graph.tex
Updated arxiv version of paper.

\subsection{PPR clustering}
\label{subsec:ppr_clustering}
We start by reviewing the PPR clustering algorithm. Let $G = (V,E)$ be an undirected, unweighted, and connected graph. We denote by $A \in \Reals^{n \times n}$ the adjacency
matrix of $G$, with entries $A_{uv} = 1$ if $(u,v) \in E$ and $0$ otherwise.  We
also denote by $D$ the diagonal degree matrix, with entries $D_{uu} :=
\sum_{v \in V} A_{uv}$, and by $I$ the $n \times n$ identity matrix. The PPR vector $p_v = p(v,\alpha;G)$ is defined with respect to a given seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, as the solution of the following linear system:
\begin{equation}
\label{eqn:ppr_vector}
p_v = \alpha e_{v} + (1 - \alpha) p_v W,
\end{equation}
where $W = (I + D^{-1}A)/2$ is the lazy random walk matrix over
$G$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in
position $v$ and 0 elsewhere).

In practice, exactly solving the system of equations~\eqref{eqn:ppr_vector} to compute the PPR vector may be too computationally expensive. To address this limitation, \citet{andersen2006} introduced the \emph{$\varepsilon$-approximate} PPR vector (aPPR), which we will denote by \smash{$p_v^{(\varepsilon)}$}. We refer the curious reader to \citet{andersen2006} for a formal algorithmic definition of the aPPR vector, and limit ourselves to highlighting a few salient points: the aPPR vector can be computed in order $\mathcal{O}(1/(\varepsilon\alpha))$ time, while satisfying the following uniform error bound: 
\begin{equation}
\label{eqn:appr_error}
\textrm{for all $u \in V$}, \quad p_v(u) - \varepsilon D_{uu}\leq
p_v^{(\varepsilon)}(u) \leq p_v(u).  
\end{equation}

Once $p_v$ or $p_v^{(\varepsilon)}$ is computed, the cluster estimate $\wh{C}$ is chosen by taking a particular sweep cut. For a given level $\beta > 0$, the \emph{$\beta$-sweep cut} of $p_v = (p_v(u))_{u \in V}$ is 
\begin{equation}
\label{eqn:sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{p_v(u)}{D_{uu}} > \beta}.
\end{equation}
To determine $\wh{C}$, one computes $S_{\beta,v}$ over all \smash{$\beta \in (L, U)$} (where the range $(L,U)$ is user-specified), and then outputs the cluster estimate
\smash{$\wh{C} = S_{\beta^*,v}$} with minimum normalized cut. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation}

\subsection{Worst-case guarantees for PPR clustering}
As mentioned, to date most analysis of local clustering has focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR, showing that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut $\Phi(\wh{C};G)$ is upper bounded by (a constant times)  $\sqrt{\Phi(C;G)}$. \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(G) := \min_{S \subseteq V} \Phi(S;G),
\end{equation}
and show that if $\Phi(C;G)$ is much smaller than $\Psi(G[C])^2$---where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$--- then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis builds on that of~\cite{zhu2013}, and we give a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}.
For now, we merely reiterate that the conclusions of \citet{andersen2006,zhu2013} cannot be straightforwardly applied to our statistical setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\mathbb{P}$, the graph $G$ is a random neighborhood graph formed by the user, and the candidate cluster is a set $\mc{C} \subset \Rd$.\footnote{Throughout, we use calligraphic notation to refer to subsets of $\Rd$.}

\subsection{PPR on a neighborhood graph}
\label{subsec:ppr_neighborhood_graph}
We now formally describe the statistical setting in which we operate, as well as the method we will study: PPR on a neighborhood graph. Let $X = \{x_1,\ldots, x_n\}$ be samples drawn i.i.d.\ from a distribution
$\Pbb$ on $\Rd$. We will assume throughout that $\Pbb$ has a density $f$ with respect to the Lebesgue measure $\nu$ on $\Rd$ . For a radius $r > 0$, we define
$G_{n,r}=(V,E)$ to be the \emph{$r$-neighborhood graph} of $X$, an
unweighted, undirected graph with vertices $V=X$, and an edge $(x_i,x_j) \in
E$ if and only if $i \neq j$ and $\|x_i - x_j\| \leq r$, where $\|\cdot\|$ is the
Euclidean norm. Once the neighborhood graph $G_{n,r}$ is formed, the PPR vector $p_v$ is then computed over $G_{n,r}$, with a resulting cluster estimate $\wh{C} \subseteq X$. The precise PPR algorithm we analyze is summarized in Algorithm~\ref{alg:ppr}.

\begin{algorithm}
	\caption{PPR on a neighborhood graph}
	\label{alg:ppr}	
	{\bfseries Input:} data $X=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
	parameter $\alpha \in [0,1]$, seed $v \in X$, sweep cut range $(L,U)$. \\     
	{\bfseries Output:} cluster estimate $\wh{C} \subseteq V$.
	\begin{algorithmic}[1]
		\STATE Form the neighborhood graph $G_{n,r}$.
		\STATE Compute the PPR vector $p_v=p(v, \alpha; G_{n,r})$ as in
		\eqref{eqn:ppr_vector}.  
		\STATE For \smash{$\beta \in (L,U)$}, compute sweep cuts $S_{\beta}$ as in
		\eqref{eqn:sweep_cuts}.\footnotemark
		\STATE Return the cluster \smash{$\wh{C} = S_{\beta^*}$}, where  
		$$
		\beta^* = \argmin_{\beta \in (L,U)}~ \Phi(S_{\beta}; G_{n,r}).
		$$
	\end{algorithmic}
\end{algorithm}
\footnotetext{Technically speaking, for each $\beta \in (L,U) \cap \{p_v(u)/D_{uu}: u \in V\}$.}

\subsection{Cluster accuracy}
We need a metric to assess the accuracy with which $\wh{C}$ estimates the candidate cluster $\mc{C}$. One commonly used metric is the misclassification error, i.e. the size of the symmetric set difference
between $\wh{C}$ and the empirical cluster $\mc{C}[X] = \mc{C} \cap X$ \citep{korostelev1993,polonik1995,rigollet2009}. We
will consider a related metric, the volume of the symmetric set difference,
which weights misclassified points according to their degree in $G_{n,r}$. To keep things simple, for a given set $S \subseteq X$ we write $\vol_{n,r}(S) := \vol(S;G_{n,r})$. 
\begin{definition}
	\label{def:volume_symmetric_set_difference}
	For an estimator \smash{$\wh{C} \subseteq X$} and a set 
	$\mathcal{C} \subseteq \Rd$, their symmetric set difference is 
	\begin{equation*}
	\wh{C} \vartriangle \mathcal{C}[X] :=
	\bigl(\wh{C} \setminus \mathcal{C}[X]\bigr) \cup
	\bigl(\mathcal{C}[X] \setminus \wh{C}\bigr).
	\end{equation*}
	Furthermore, we denote the volume of the symmetric set difference by 
	$$
	\Delta(\wh{C}, \mathcal{C}[X]) := \vol_{n,r}(\wh{C} \vartriangle \mathcal{C}[X]). 
	$$
\end{definition}

\subsection{Population normalized cut, conductance, and local spread}
Next we define three population-level functionals of $\mc{C}$---the normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and local spread $s_{\Pbb,r}(\mc{C})$---which govern the volume of the symmetric set difference $\Delta(\wh{C},\mc{C}[X])$. Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e. let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$. Also let $\deg_{\Pbb,r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\Pbb(y)$ to be the expected degree of $x$ in $G_{n,r}$. 
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

Let $\wt{\mathbb{P}}(\cdot) = \mathbb{P}(\cdot|x \in \mc{C})$ be the conditional distribution of $x$, i.e let $\wt{\mbb{P}}(\mc{S}) = \wt{\Pbb}(\mc{S} \cap \mc{C})/\wt{\Pbb}(\mc{C})$ for measurable sets $\mc{S} \subseteq \Rd$.

\begin{definition}[Population conductance]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population conductance} is
	\begin{equation}
	\label{eqn:population_conductance}
	\Psi_{\mathbb{P},r}(\mc{C}) = \inf_{\mc{S} \subseteq \mc{C}} \Phi_{\wt{\Pbb},r}(\mc{S}).
	\end{equation}
\end{definition}
\begin{definition}[Population local spread]
	For a set $\mc{C} \subset \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population local spread} is
	\begin{equation}
	\label{eqn:local_spread}
	s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
	\end{equation}
\end{definition}

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, by construction these functionals are quite obviously the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities in turn suffice to upper bound the volume of the symmetric set difference. For this reason, similar population level functionals are used by \citep{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}.

@@

\subsection{Main Results}
We now informally state our two main upper bounds, regarding the recovery of a generic cluster $\mc{C}$, and a density cluster $\mc{C}_{\lambda}$. Theorem~\ref{thm:volume_ssd_ub_informal} informally summarizes the first of our main results (formally stated in Theorem~\ref{thm:volume_ssd_ub}) regarding the recovery of a generic cluster $\mc{C}$.
\begin{theorem}[Informal]
	\label{thm:volume_ssd_ub_informal}
	Let $\mc{C} \subset \Rd$ and $\Pbb$ satisfy appropriate regularity conditions, and suppose Algorithm~\ref{alg:ppr} is well-initialized with respect to $\mc{C}$. For all $n \in \mathbb{N}$ sufficiently large, with high probability it holds that
	\begin{equation*}
	\frac{\Delta(\wh{C},\mc{C}[X])}{\vol_{n,r}(\mc{C}[X])} \leq c \cdot\Phi_{\Pbb,r}(\mc{C}) \cdot  \biggl(\frac{\log(1/s_{\Pbb,r}(\mc{C}))}{\Psi_{\Pbb,r}(\mc{C})}\biggr)^2.
	\end{equation*}
\end{theorem}
(In the above, and throughout, $c$ stands for a universal constant that may change from line to line.) Put even more succinctly, we find that $\Delta(\wh{C},\mc{C}[X])$ is small when $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $\bigl(\Psi_{\Pbb,r}(\mc{C})/\log(1/s_{\Pbb,r}(\mc{C}))\bigr)^2$. To the best of our knowledge, this gives the first population-level guarantees for local clustering in the nonparametric statistical context.

>>>>

\subsection{PPR Clustering}
\label{subsec:ppr_clustering}
We start by reviewing the PPR clustering algorithm. Let $G = (V,E)$ be an undirected, unweighted, and connected graph. We denote by $A \in \Reals^{n \times n}$ the adjacency matrix of $G$, with entries $A_{uv} = 1$ if $(u,v) \in E$ and $0$ otherwise.  We also denote by $D$ the diagonal degree matrix, with entries $D_{uu} := \sum_{v \in V} A_{uv}$, and by $I$ the $n \times n$ identity matrix. The PPR vector $p_v = p(v,\alpha;G)$ is defined with respect to a given seed node $v \in V$ and a teleportation parameter $\alpha \in [0,1]$, as the solution of the following linear system:
\begin{equation}
\label{eqn:ppr_vector}
p_v = \alpha e_{v} + (1 - \alpha) p_v W,
\end{equation}
where $W = (I + D^{-1}A)/2$ is the lazy random walk matrix over $G$ and $e_{v}$ is the indicator vector for node $v$ (that has a 1 in position $v$ and 0 elsewhere).

In practice, exactly solving the system of equations~\eqref{eqn:ppr_vector} to compute the PPR vector may be too computationally expensive. To address this limitation, \citet{andersen2006} introduced the \emph{$\varepsilon$-approximate} PPR vector (aPPR), which we will denote by \smash{$p_v^{(\varepsilon)}$}. We refer the curious reader to \citet{andersen2006} for a formal algorithmic definition of the aPPR vector, and limit ourselves to highlighting a few salient points: the aPPR vector can be computed in order $\mc{O}(1/(\varepsilon\alpha))$ time, while satisfying the following uniform error bound: 
\begin{equation}
\label{eqn:appr_error}
p_v(u) - \varepsilon D_{uu}\leq p_v^{(\varepsilon)}(u) \leq p_v(u), 
\quad \text{for all $u \in V$}.
\end{equation}

Once $p_v$ or \smash{$p_v^{(\varepsilon)}$} is computed, the cluster estimate \smash{$\wh{C}$} is chosen by taking a particular sweep cut. For a given level $\beta > 0$, the \emph{$\beta$-sweep cut} of $p_v = (p_v(u))_{u \in V}$ is 
\begin{equation}
\label{eqn:sweep_cuts}
S_{\beta,v} := \set{u \in V: \frac{p_v(u)}{D_{uu}} > \beta}.
\end{equation}
To define \smash{$\wh{C}$}, one computes $S_{\beta,v}$ over all \smash{$\beta \in (L, U)$} (where the range $(L,U)$ is user-specified), and then outputs the cluster estimate \smash{$\wh{C} = S_{\beta^*,v}$} with minimum normalized cut. For a set $C \subseteq V$ with complement $C^c = V \!\setminus\! C$, the \emph{cut} and \emph{volume} are respectively,
\begin{equation}
\label{eqn:cut_volume}
\cut(C;G) := \sum_{u \in C} \sum_{v \in C^c}
\1\{(u,v) \in E\},~~ \vol(C; G) := \sum_{u \in C}  \sum_{v \in V} \1\{(u,v) \in E\},
\end{equation}
and the \emph{normalized cut} of $C$ is
\begin{equation}
\label{eqn:normalized_cut}
\Phi(C; G) := \frac{\cut(C;G)}{\min \set{\vol(C; G), \vol(C^c; G)}}.
\end{equation}

\subsection{Worst-Case Guarantees for PPR Clustering}
As mentioned, most analyses of local clustering have focused on worst-case guarantees, defined with respect to functionals of an a priori fixed graph $G = (V,E)$.  For instance, \cite{andersen2006} analyze the normalized cut of the cluster estimate $\wh{C}$ output by PPR, showing that when PPR is appropriately seeded within a candidate cluster $C \subseteq V$, the normalized cut $\Phi(\wh{C};G)$ is upper bounded by (a constant times)  $\sqrt{\Phi(C;G)}$. \cite{zhu2013} build on this: they introduce a second functional, the \emph{conductance} $\Psi(G)$, defined as 
\begin{equation}
\label{eqn:conductance}
\Psi(G) := \min_{S \subseteq V} \Phi(S;G),
\end{equation}
and show that if $\Phi(C;G)$ is much smaller than $\Psi(G[C])^2$---where $G[C] = (C,E \cap (C \times C))$ is the subgraph of $G$ induced by $C$--- then (in addition to having a small normalized cut) the cluster estimate $\wh{C}$ approximately recovers $C$. Our own analysis builds on that of~\cite{zhu2013}, and we present a more detailed summary of their results in Section~\ref{sec:ub_symmetric_set_difference}. For now, we merely reiterate that the conclusions of \citet{andersen2006,zhu2013} cannot be straightforwardly applied to our setting, where the input data are random samples $\{x_1,\ldots,x_n\}$ drawn from a distribution $\Pbb$, the graph $G$ is a random neighborhood graph formed from the samples, and the candidate cluster is a set $\mc{C} \subseteq \Rd$.\footnote{Throughout, we use calligraphic notation to refer to subsets of $\Rd$.}

\subsection{PPR on a Neighborhood Graph}
\label{subsec:ppr_neighborhood_graph}
We now formally describe the statistical setting in which we operate, as well as the method we will study: PPR on a neighborhood graph. Let $X = \{x_1,\ldots, x_n\}$ be samples drawn i.i.d.\ from a distribution $\Pbb$ on $\Rd$. We will assume throughout that $\Pbb$ has a density $f$ with respect to the Lebesgue measure $\nu$ on $\Rd$ . For a radius $r > 0$, we define $G_{n,r}=(V,E)$ to be the \emph{$r$-neighborhood graph} of $X$, an unweighted, undirected graph with vertices $V=X$, and an edge $(x_i,x_j) \in E$ if and only if $i \neq j$ and $\|x_i - x_j\| \leq r$, where $\|\cdot\|$ is the Euclidean norm. Once the neighborhood graph $G_{n,r}$ is formed, the PPR vector $p_v$ is then computed over $G_{n,r}$, with a resulting cluster estimate \smash{$\wh{C} \subseteq X$}. The precise algorithm is summarized in Algorithm~\ref{alg:ppr}.

\begin{algorithm}
	\caption{PPR on a neighborhood graph}
	\label{alg:ppr}	
	{\bfseries Input:} data $X=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation
	parameter $\alpha \in [0,1]$, seed $v \in X$, sweep cut range $(L,U)$. \\     
	{\bfseries Output:} cluster estimate $\wh{C} \subseteq V$.
	\begin{algorithmic}[1]
		\STATE Form the neighborhood graph $G_{n,r}$.
		\STATE Compute the PPR vector $p_v=p(v, \alpha; G_{n,r})$ as in
		\eqref{eqn:ppr_vector}.  
		\STATE Compute sweep cuts $S_{\beta}$ as in \eqref{eqn:sweep_cuts}, for each $\beta \in (L,U)$.\footnotemark 
		\STATE Return \smash{$\wh{C} = S_{\beta^*}$}, where  
		$$
		\beta^* = \argmin_{\beta \in (L,U)}~ \Phi(S_{\beta}; G_{n,r}).
		$$
	\end{algorithmic}
\end{algorithm}
\footnotetext{Technically speaking, for each $\beta \in (L,U) \cap \{p_v(u)/D_{uu}: u \in V\}$.}

\subsection{Cluster Accuracy}
We need a metric to assess the accuracy with which $\wh{C}$ estimates the candidate cluster $\mc{C}$. One commonly used metric is the misclassification error, i.e., the size of the symmetric set difference between \smash{$\wh{C}$} and the empirical cluster $\mc{C}[X] = \mc{C} \cap X$ \citep{korostelev1993,polonik1995,rigollet2009}. We will consider a related metric, the volume of the symmetric set difference, which weights misclassified points according to their degree in $G_{n,r}$. To keep things simple, for a given set $S \subseteq X$ we write $\vol_{n,r}(S) := \vol(S;G_{n,r})$.

\begin{definition}
	\label{def:volume_symmetric_set_difference}
	For an estimator \smash{$\wh{C} \subseteq X$} and a set $\mc{C} \subseteq \Rd$, their symmetric set difference is  
	\begin{equation*}
	\wh{C} \vartriangle \mc{C}[X] :=
	\bigl(\wh{C} \setminus \mc{C}[X]\bigr) \cup
	\bigl(\mc{C}[X] \setminus \wh{C}\bigr).
	\end{equation*}
	Furthermore, we denote the volume of the symmetric set difference by 
	$$
	\Delta(\wh{C}, \mc{C}[X]) := \vol_{n,r}(\wh{C} \vartriangle \mc{C}[X]). 
	$$
\end{definition}

\subsection{Population Normalized Cut, Conductance, and Local Spread}
Next we define three population-level functionals of $\mc{C}$---the normalized cut $\Phi_{\Pbb,r}(\mc{C})$, conductance $\Psi_{\Pbb,r}(\mc{C})$, and the local spread $s_{\Pbb,r}(\mc{C})$---which we will use to upper bound the volume of the symmetric set difference \smash{$\Delta(\wh{C},\mc{C}[X])$} with high probability.

Let the population-level \emph{cut} of $\mc{C}$ be the expectation (up to a rescaling) of $\cut_{n,r}(\mc{C}[X]) := \cut(\mc{C}[X]; G_{n,r})$,  and likewise let the population-level \emph{volume} of $\mc{C}$ be the expectation (up to a rescaling) of $\vol_{n,r}(\mc{C}[X]) := \vol(\mc{C}[X]; G_{n,r})$; i.e., let
\begin{equation*}
\mathrm{cut}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\mc{C}^c} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),~~ \mathrm{vol}_{\Pbb,r}(\mc{C}) := \int_{\mc{C}} \int_{\Rd} \1\{\|x - y\| \leq r\} \,d\Pbb(y) \,d\Pbb(x),
\end{equation*}
where $\mc{C}^c := \Rd \!\setminus\! \mc{C}$. Also let $\deg_{\Pbb,r}(x) := \int_{\Rd} \1\{\|y - x\| \leq r\} \,d\Pbb(y)$ be the expected degree of $x$ in $G_{n,r}$. 
\begin{definition}[Population normalized cut]
	For a set $\mc{C} \subseteq \Rd$, distribution $\Pbb$, and radius $r > 0$, the \emph{population normalized cut} is
	\begin{equation}
	\label{eqn:population_normalized_cut}
	\Phi_{\Pbb,r}(\mc{C}) := \frac{\mathrm{cut}_{\Pbb,r}(\mc{C})}{\min\{\mathrm{vol}_{\Pbb,r}(\mc{C}), \mathrm{vol}_{\Pbb,r}(\mc{C}^c)\}}.
	\end{equation}
\end{definition}

Let \smash{$\wt{\Pbb}(\cdot) = \Pbb(\cdot|x \in \mc{C})$} be the conditional distribution of $x$, i.e., let \smash{$\wt{\Pbb}(\mc{S}) = \wt{\Pbb}(\mc{S} \cap \mc{C})/\wt{\Pbb}(\mc{C})$} for measurable sets $\mc{S} \subseteq \Rd$.

\begin{definition}[Population conductance]
	For a set $\mc{C} \subseteq \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population conductance} is
	\begin{equation}
	\label{eqn:population_conductance}
	\Psi_{\Pbb,r}(\mc{C}) = \inf_{\mc{S} \subseteq \mc{C}} \Phi_{\wt{\Pbb},r}(\mc{S}).
	\end{equation}
\end{definition}

\begin{definition}[Population local spread]
	For a set $\mc{C} \subseteq \Rd$, distribution $\Pbb$ and radius $r > 0$, the \emph{population local spread} is
	\begin{equation}
	\label{eqn:local_spread}
	s_{\Pbb,r}(\mc{C}) := \min_{x \in \mc{C}} \biggl\{\frac{\bigl(\deg_{\wt{\Pbb},r}(x)\bigr)^2}{\vol_{\wt{\Pbb},r}(\mc{C})} \biggr\},
	\end{equation}
\end{definition}

It is quite natural that $\Phi_{\Pbb,r}(\mc{C})$ and $\Psi_{\Pbb,r}(\mc{C})$ should help quantify the role geometry plays in local spectral clustering. Indeed, these functionals are the population-level analogues of the empirical quantities $\Phi_{n,r}(\mc{C}[X]) := \Phi(\mc{C}[X];G_{n,r})$ and $\Psi_{n,r}(\mc{C}[X]) := \Psi(G_{n,r}\bigl[\mc{C}[X]\bigr])$, and as we have already mentioned, these empirical quantities can be used to upper bound the volume of the symmetric set difference. For this reason, similar population-level functionals are used by \cite{shi2009,schiebinger2015,garciatrillos19} in the analysis of \emph{global} spectral clustering in a statistical context. We will comment more on the relationship between these works and our own results in Section~\ref{subsec:related_work}.

@@

\subsection{Main Results}
We now informally state our two main upper bounds, regarding the recovery of a generic cluster $\mc{C}$, and a density cluster $\mc{C}_{\lambda}$. Theorem~\ref{thm:volume_ssd_ub_informal} informally summarizes the first of our main results (formally stated in Theorem~\ref{thm:volume_ssd_ub}) regarding the recovery of a generic cluster $\mc{C}$.
\begin{theorem}[Informal]
	\label{thm:volume_ssd_ub_informal}
	If $\mc{C} \subseteq \Rd$ and $\Pbb$ satisfy appropriate regularity conditions, and Algorithm~\ref{alg:ppr} is initialized properly with respect to $\mc{C}$, then for all sufficiently large $n$, with high probability it holds that
	\begin{equation*}
	\frac{\Delta(\wh{C},\mc{C}[X])}{\vol_{n,r}(\mc{C}[X])} \leq c \cdot\Phi_{\Pbb,r}(\mc{C}) \cdot  \biggl(\frac{\log(1/s_{\Pbb,r}(\mc{C}))}{\Psi_{\Pbb,r}(\mc{C})}\biggr)^2.
	\end{equation*}
\end{theorem}

(Above, and throughout, $c$ stands for a universal constant that may change from line to line.) Put more succinctly, we find that \smash{$\Delta(\wh{C},\mc{C}[X])$} is small when $\Phi_{\Pbb,r}(\mc{C})$ is small relative to $(\Psi_{\Pbb,r}(\mc{C})/\log(1/s_{\Pbb,r}(\mc{C})))^2$. To the best of our knowledge, this gives the first population-level guarantees for local clustering in the nonparametric statistical context.

Next, Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} informally summarizes the second of our main results (formally stated in Theorem~\ref{thm:density_cluster_volume_ssd_ub}) regarding the recovery of a $\lambda$-density cluster $\mc{C}_{\lambda}$ by PPR. For reasons that we explain later in Section~\ref{sec:ppr_density_cluster}, our cluster recovery statement will actually be with respect to the $\sigma$-thickened set $\mc{C}_{\lambda,\sigma} := \{x \in \Rd: \mathrm{dist}(x,\mc{C}_{\lambda}) < \sigma\}$, for a given $\sigma > 0$. The upper bound we establish is a function of various parameters that measure the conditioning of both the density cluster $\mc{C}_{\lambda,\sigma}$ and density $f$ for recovery by PPR. We assume that $\mc{C}_{\lambda,\sigma}$ is the image of a convex set $\mc{K}$ of finite diameter $\mathrm{diam}(\mc{K}) \leq \rho < \infty$ under a Lipschitz, measure-preserving mapping $g$, with Lipschitz constant $M$. We also assume that $f$ is bounded away from $0$ and $\infty$ on $\mc{C}_{\lambda,\sigma}$:
\begin{equation*}
0 < \lambda_{\sigma} \leq f(x) \leq \Lambda_{\sigma} < \infty~~\text{for all $x \in \mc{C}_{\lambda,\sigma}$},
\end{equation*}
and additionally satisfies the following low-noise condition:
\begin{equation*}
\inf_{y \in \mc{C}_{\lambda,\sigma}} f(y) - f(x) \geq  \theta \cdot \dist(x, \mc{C}_{\lambda,\sigma})^{\gamma}~~\text{for all $x$ such that $0 < \dist(x,\mc{C}_{\lambda,\sigma}) \leq \sigma$.}
\end{equation*}
(Here $\dist(x,\mc{C}) := \inf_{y \in \mc{C}} \|y - x\|$.) 
\begin{theorem}[Informal]
	\label{thm:density_cluster_volume_ssd_ub_informal}
	If $\mc{C}_{\lambda} \subseteq \Rd$ is a $\lambda$-density cluster of a distribution $\Pbb$, which satisfies appropriate regularity conditions, and Algorithm~\ref{alg:ppr} is initialized properly with respect to $\mc{C}_{\lambda,\sigma}$, then for all sufficiently large $n$, with high probability it holds that
	\begin{equation*}
	\frac{\Delta(\wh{C},\mc{C}_{\lambda,\sigma}[X])}{\vol_{n,r}(\mc{C}_{\lambda,\sigma})} \leq c \cdot d^4 \cdot \frac{M^2\rho^2}{\sigma r} \cdot \frac{\Lambda_{\sigma}^2 \lambda (\lambda - \theta \frac{r^{\gamma}}{\gamma + 1})}{\lambda_{\sigma}^4} \cdot \log^2\biggl(\frac{\Lambda_{\sigma}^{2/d} M\rho}{\lambda_{\sigma}^{2/d}2r}\biggr).
	\end{equation*}
\end{theorem}

The above result reveals the separate roles played by geometry and density in the ability of PPR to recover a density cluster. Here $M$, $\rho$, and $\sigma$ capture whether $\mc{C}_{\lambda,\sigma}$ is geometrically well-conditioned (short and fat) or poorly-conditioned (long and thin) for recovery by PPR. Likewise, the parameters $\lambda_{\sigma}, \Lambda_{\sigma}, \gamma$, and $\theta$ measure whether the density $f$ is well-conditioned (approximately uniform over the density cluster, and having thin tails outside of it) or poorly conditioned (vice versa). Theorem~\ref{thm:density_cluster_volume_ssd_ub_informal} says that if the thickened density cluster $\mc{C}_{\lambda,\sigma}$ is geometrically well-conditioned---meaning, $M^2\rho^2/(\sigma r) \approx 1$---and the density $f$ is well-conditioned near $\mc{C}_{\lambda,\sigma}$---meaning, $\Lambda_{\sigma} \approx \lambda \approx \lambda_{\sigma}$ and $\lambda - \theta r^{\gamma}/(\gamma + 1)$ is much less than $\lambda_{\sigma}$---then PPR will approximately recover $\mc{C}_{\lambda,\sigma}$.

\subsection{Related Work}
\label{subsec:related_work}
We now summarize some related work (in addition to the background material already given above), regarding the theory of spectral clustering, and of density cluster recovery.

\subsubsection{Spectral Clustering} 
In the stochastic block model (SBM), arguably one of the simplest models of network formation, edges between nodes independently occur with probability based on a latent community membership. In the SBM, the ability of spectral algorithms to perform clustering---or community detection---is well-understood, dating back to \citet{mcsherry2001} who gives conditions under which the entire community structure can be recovered. In more recent work, \citet{rohe2011} upper bound the fraction of nodes misclassified by a spectral algorithm for the high-dimensional (large number of blocks) SBM, and \citet{lei2015} extend these results to the sparse (low average degree) regime. Relatedly, \citet{clauset08,balakrishnan2011,li2018} analyze the misclassification rate when the block model exhibits some hierarchical structure. The framework we consider, in which nodes correspond to data points sampled from an underlying density, and edges between nodes are formed based on geometric proximity, is quite different than the SBM, and therefore so is our analysis.
